---
title: "platformbias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{platformbias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The setup is that there are two datasets, each with potentially several ancestries, and the data are allele counts for each dataset, ancestry, and at several loci.
The goal of `af_test` is to determine, separately for each locus, whether the allele frequencies are essentially the same in both datasets or not, in a way that combines the data of all ancestries.
The null hypothesis is that all allele frequencies, which can be different for each ancestry, are the same across studies.
The alternative hypothesis is that at least one allele frequency is different in at least one ancestry.
Thus, the alternative has $K$ degrees of freedom compared to the null, where $K$ is the number of ancestries.
This test does not measure or care about the direction of differences, in fact in one ancestry the allele frequency can be smaller in dataset 1 vs 2, but in another ancestry it can be larger, and these count as overall differences (they do not cancel each other out).
Thus, this test is appropriate if the alternative hypothesis does not have a clear direction or structure, we just want to know if there are differences, but may be underpowered when there is a consistent direction or other structure in the data.

There is a second function, `af_test_or`, that also calculates odds ratios (ORs) in a more association-like test, which assumes a more restricted alternative hypothesis that the OR is the same across ancestries.
Each ancestry can still have a different allele frequency under the null.
The null hypothesis is the same for both functions, here stated as the OR=1 (for all ancestries).
Thus, the alternative hypothesis has a single degree of freedom, and the test will have more power when this assumption is correct, which is expected for disease causing variants, for example.
Furthermore, `af_test_or` is much more computationally intensive than `af_test`.

Below we simulate data, to both illustrate usage and to evaluate differences in the behavior of the two statistical tests.

## Allele frequency testing of separate loci, single ancestry

First we consider the simplest setting, where there is a single ancestry, so the two tests are expected to agree in their alternative hypothesis (they only differ when there are multiple ancestries).
We will use allele counts to perform these statistical tests while taking sample sizes into account.

```{r setup}
library(platformbias)
```

### Null data: equal allele frequency in both datasets

Let's simulate a large number of "null" SNPs for the first example where allele frequencies are truly the same.
Let's simulate the desired data using a compact vectorized notation.
```{r}
# number of loci to simulate
m <- 100
# this is bonferroni threshold for significance
p_cut_bonferroni <- 0.05 / m
p_cut_bonferroni

# draw `m` random allele frequencies between 0 and 1 from Uniform distribution.
# (test is applied to each locus separately, so it does not assume any particular distribution
# for the true allele frequencies).
p <- runif( m )

# for fun, draw random sample sizes for each locus and for each of the two datasets.
# use the Poisson distribution to draw `m` random non-negative integers with mean `n_mean`.
# don't allow zero sample sizes though, minimum is 1.
# (as above, there is no assumption being made for sample size distributions,
# each locus is tested separately and sample sizes are known and conditioned upon.)
n_mean <- 1000
n1 <- 1 + rpois( m, n_mean )
n2 <- 1 + rpois( m, n_mean )

# finally, draw the observed counts, a vector of length `m`, each element a count
# between zero and `n1` and mean `n1*p`, both of which varies per locus.
# Use Binomial distribution, which **is** assumed by the statistical test.
x1 <- rbinom( m, n1, p )
x2 <- rbinom( m, n2, p )

# apply test to each locus separately
obj1 <- af_test( x1, n1, x2, n2 )
pvalues1 <- obj1$pval

# use second function now, which accepts the data in exactly the same format, but the outputs are different
obj2 <- af_test_or( x1, n1, x2, n2 )
pvalues2 <- obj2$pval
```

As the data was simulated under the null distribution (the true allele frequencies in both datasets are the same), we expect these p-values (for both functions) to be uniformly distributed, whose density height is marked by the red line below.
Let's confirm:
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center'}
par( mfrow = c(1, 2) )
hist( pvalues1, freq = FALSE, main = 'af_test' )
abline( h = 1, lty = 2, col = 'red' )
hist( pvalues2, freq = FALSE, main = 'af_test_or' )
abline( h = 1, lty = 2, col = 'red' )
```
The p-values of both functions are not identical, but they agree very well, as expected!
Also as expected, nothing was significant by the Bonferroni criterion (red lines).
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center'}
# set a nice joint range that always includes significance threshold
lims <- range( pvalues1, pvalues2, p_cut_bonferroni )
plot( pvalues1, pvalues2, log = 'xy', xlim = lims, ylim = lims )
# mark y=x line
abline( 0, 1, lty = 2, col = 'gray' )
# mark significance threshold
abline( h = p_cut_bonferroni, lty = 2, col = 'red' )
abline( v = p_cut_bonferroni, lty = 2, col = 'red' )
```
For the second function, let's look at the estimated ORs and CIs.
Here the true OR is 1, and we can see that it is contained in the CIs most of the time, as desired.
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center', warning = FALSE}
library(plotrix)
# hack limits because we do get infinities
# sample sizes are small too so CIs can be very wide
# always ignore zeroes because of log scale
ylim <- c(
    quantile( obj2$CIL[ obj2$CIL > 0 ], 0.05, na.rm = TRUE ),
    quantile( obj2$CIU[ obj2$CIU > 0 ], 0.95, na.rm = TRUE )
)
plotCI(
    x = 1 : m,
    y = obj2$OR,
    li = obj2$CIL,
    ui = obj2$CIU,
    ylim = ylim,
    log = 'y',
    xlab = 'Locus',
    ylab = 'Estimated OR (95% CI)'
)
abline( h = 1, lty = 2, col = 'gray' )
```

### Alternative data: different allele frequencies in both datasets

Now let's create data where the alternative hypothesis holds, namely that the allele frequencies are different in the each dataset.
Let's simply create new allele frequencies for the second dataset, regenerate its data, and retest.
These p-values are as expected, namely non-uniform and concentrated around zero.
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center'}
# draw new allele frequencies for second dataset
p2 <- runif( m )
# and new allele counts from these frequencies
x2 <- rbinom( m, n2, p2 )

# apply tests to each locus separately
obj1 <- af_test( x1, n1, x2, n2 )
pvalues1 <- obj1$pval
obj2 <- af_test_or( x1, n1, x2, n2 )
pvalues2 <- obj2$pval

# plot new p-value distributions
par( mfrow = c(1, 2) )
hist( pvalues1, freq = FALSE, main = 'af_test' )
abline( h = 1, lty = 2, col = 'red' )
hist( pvalues2, freq = FALSE, main = 'af_test_or' )
abline( h = 1, lty = 2, col = 'red' )
```
The p-values of both functions agree to some extent, often declaring the same loci significant, but there is more power (smaller p-values) for `af_test`:
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center'}
lims <- range( pvalues1, pvalues2, p_cut_bonferroni )
plot( pvalues1, pvalues2, log = 'xy', xlim = lims, ylim = lims )
abline( 0, 1, lty = 2, col = 'gray' )
abline( h = p_cut_bonferroni, lty = 2, col = 'red' )
abline( v = p_cut_bonferroni, lty = 2, col = 'red' )
```
In this case the true ORs vary per locus, so let's compare them to estimates directly, seeing good unbiased estimates!
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center', warning = FALSE}
library(plotrix)
true_OR <- p * ( 1 - p2 ) / ( p2 * ( 1 - p ) )
# always ignore zeroes because of log scale
xlim <- range( true_OR[ true_OR > 0 ] )
ylim <- c(
    quantile( obj2$CIL[ obj2$CIL > 0 ], 0.05, na.rm = TRUE ),
    quantile( obj2$CIU[ obj2$CIU > 0 ], 0.95, na.rm = TRUE )
)
plotCI(
    x = true_OR,
    y = obj2$OR,
    li = obj2$CIL,
    ui = obj2$CIU,
    xlim = xlim,
    ylim = ylim,
    log = 'xy',
    xlab = 'True OR',
    ylab = 'Estimated OR (95% CI)'
)
abline( 0, 1, lty = 2, col = 'gray' )
```


## Allele frequency testing of separate loci, combining data from different ancestries

The same functions accept matrix inputs, in which case column data is combined in a single test, but rows remain separate tests.
The matrix dimensions of all inputs must match.
The `af_test` test has degrees of freedom equal to the number of columns, and tests the joint null that all allele frequencies are equal (that is, even a single departure favors the alternative).
In contrast, `af_test_or` has a single degree of freedom, and tests the null that the odds ratio, which is the same across ancestries, is not equal to one.
Let's simulate more data under their null but with multiple columns now.
It's useful to think of the rows as loci and columns as different ancestries, as done below, but they may represent different groupings or categories as long as columns are combined but rows are not.

### Null data: equal allele frequency in both datasets

```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center'}
# number of ancestries to simulate
k <- 3

# draw `m*k` random allele frequencies, reshape as matrix for clarity (code works anyway)
p <- matrix( runif( m*k ), nrow = m, ncol = k )

# draw random sample sizes for each locus, ancestry, and the two datasets, as before.
n_mean <- 100
n1 <- 1 + rpois( m*k, n_mean )
n2 <- 1 + rpois( m*k, n_mean )
n1 <- matrix( n1, nrow = m, ncol = k )
n2 <- matrix( n2, nrow = m, ncol = k )

# finally, draw the observed counts
x1 <- rbinom( m*k, n1, p )
x2 <- rbinom( m*k, n2, p )
x1 <- matrix( x1, nrow = m, ncol = k )
x2 <- matrix( x2, nrow = m, ncol = k )

# apply tests to each locus separately, but combining column data
obj1 <- af_test( x1, n1, x2, n2 )
pvalues1 <- obj1$pval
obj2 <- af_test_or( x1, n1, x2, n2 )
pvalues2 <- obj2$pval
# confirm that there are exactly `m` p-values (not `k`, `m*k`, or some other number)
stopifnot( length( pvalues1 ) == m )
stopifnot( length( pvalues2 ) == m )
# confirm that they are uniform as expected
par( mfrow = c(1, 2) )
hist( pvalues1, freq = FALSE, main = 'af_test' )
abline( h = 1, lty = 2, col = 'red' )
hist( pvalues2, freq = FALSE, main = 'af_test_or' )
abline( h = 1, lty = 2, col = 'red' )
```
The p-values of both functions disagree more here, because there are multiple ancestries, though they're all insignificant.
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center'}
lims <- range( pvalues1, pvalues2, p_cut_bonferroni )
plot( pvalues1, pvalues2, log = 'xy', xlim = lims, ylim = lims )
abline( 0, 1, lty = 2, col = 'gray' )
abline( h = p_cut_bonferroni, lty = 2, col = 'red' )
abline( v = p_cut_bonferroni, lty = 2, col = 'red' )
```
Let's again look at estimated ORs, which are about 1, as expected here.
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center', warning = FALSE}
ylim <- c(
    quantile( obj2$CIL, 0.05 ),
    quantile( obj2$CIU, 0.95 )
)
plotCI(
    x = 1 : m,
    y = obj2$OR,
    li = obj2$CIL,
    ui = obj2$CIU,
    ylim = ylim,
    log = 'y',
    xlab = 'Locus',
    ylab = 'Estimated OR (95% CI)'
)
abline( h = 1, lty = 2, col = 'gray' )
```

### Alternative 1 data: different allele frequencies in both datasets, different ORs per ancestry

And, as before, create a second dataset with different allele frequencies, which yield p-values that peak near zero.
These are simulated under the null of `af_test`, where differences are in random directions for different ancestries with no consistent OR, so `af_test` should have more power than `af_test_or`.
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center', error = TRUE}
# draw `m*k` random allele frequencies
p2 <- matrix( runif( m*k ), nrow = m, ncol = k )
# redraw the observed counts
x2 <- rbinom( m*k, n2, p2 )
x2 <- matrix( x2, nrow = m, ncol = k )

# apply tests
obj1 <- af_test( x1, n1, x2, n2 )
pvalues1 <- obj1$pval
obj2 <- af_test_or( x1, n1, x2, n2 )
pvalues2 <- obj2$pval
# confirm that there are exactly `m` p-values
stopifnot( length( pvalues1 ) == m )
stopifnot( length( pvalues2 ) == m )
# confirm that they peaked near zero as expected
par( mfrow = c(1, 2) )
hist( pvalues1, freq = FALSE, main = 'af_test' )
abline( h = 1, lty = 2, col = 'red' )
hist( pvalues2, freq = FALSE, main = 'af_test_or' )
abline( h = 1, lty = 2, col = 'red' )
```
Indeed, `af_test` has smaller p-values, and more significant loci, than `af_test_or` in this evaluation:
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center'}
lims <- range( pvalues1, pvalues2, p_cut_bonferroni )
plot( pvalues1, pvalues2, log = 'xy', xlim = lims, ylim = lims )
abline( 0, 1, lty = 2, col = 'gray' )
abline( h = p_cut_bonferroni, lty = 2, col = 'red' )
abline( v = p_cut_bonferroni, lty = 2, col = 'red' )
```

For the previous alternative data, it is unclear what the overall OR should be when every ancestry potentially has a different OR.
After trying some failed experiments, I empirically saw that the OR of the average allele frequencies (averaged across ancestries) is closest to the quantity being estimated:
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center', warning = FALSE}
# average ancestries before OR
p1a <- rowMeans( p )
p2a <- rowMeans( p2 )
true_OR <- p1a * ( 1 - p2a ) / ( p2a * ( 1 - p1a ) )
ylim <- c(
    quantile( obj2$CIL, 0.05 ),
    quantile( obj2$CIU, 0.95 )
)
plotCI(
    x = true_OR,
    y = obj2$OR,
    li = obj2$CIL,
    ui = obj2$CIU,
    ylim = ylim,
    log = 'xy',
    xlab = 'True OR',
    ylab = 'Estimated OR (95% CI)'
)
abline( 0, 1, lty = 2, col = 'gray' )
```

### Alternative 2 data: different allele frequencies in both datasets, equal ORs across ancestries

Lastly, let's simulate alternative data for the `af_test_or` model, which assumes that the ORs are shared across ancestries, and for which that test ought to have higher power than `af_test` does.
For simplicity, the true OR will further be equal across loci.
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center', error = TRUE}
# choose a desired OR large enough for its CIs to not overlap 1
true_OR <- 3
# construct allele frequencies with that true OR relative to the original `p`
p2 <- p / ( p * ( 1 - true_OR ) + true_OR )
# redraw the observed counts
x2 <- rbinom( m*k, n2, p2 )
x2 <- matrix( x2, nrow = m, ncol = k )

# apply tests
obj1 <- af_test( x1, n1, x2, n2 )
pvalues1 <- obj1$pval
obj2 <- af_test_or( x1, n1, x2, n2 )
pvalues2 <- obj2$pval
# confirm that there are exactly `m` p-values
stopifnot( length( pvalues1 ) == m )
stopifnot( length( pvalues2 ) == m )
# confirm that they peaked near zero as expected
par( mfrow = c(1, 2) )
hist( pvalues1, freq = FALSE, main = 'af_test' )
abline( h = 1, lty = 2, col = 'red' )
hist( pvalues2, freq = FALSE, main = 'af_test_or' )
abline( h = 1, lty = 2, col = 'red' )
```
As expected, `af_test_or` has smaller p-values than `af_test` in this setting, though the difference is small and the number of significant loci similar both ways:
```{r, fig.width = 4.2, fig.height = 4.2, fig.align = 'center'}
lims <- range( pvalues1, pvalues2, p_cut_bonferroni )
plot( pvalues1, pvalues2, log = 'xy', xlim = lims, ylim = lims )
abline( 0, 1, lty = 2, col = 'gray' )
abline( h = p_cut_bonferroni, lty = 2, col = 'red' )
abline( v = p_cut_bonferroni, lty = 2, col = 'red' )
```
The estimated ORs and CIs look great here, most of the time confidently excluding the null value OR=1!
```{r, fig.width = 6.5, fig.height = 3, fig.align = 'center', warning = FALSE}
ylim <- c(
    min( quantile( obj2$CIL, 0.05 ), 1 ),
    quantile( obj2$CIU, 0.95 )
)
plotCI(
    x = 1 : m,
    y = obj2$OR,
    li = obj2$CIL,
    ui = obj2$CIU,
    ylim = ylim,
    log = 'y',
    xlab = 'Locus',
    ylab = 'Estimated OR (95% CI)'
)
# desired value
abline( h = true_OR, lty = 2, col = 'red' )
# null value
abline( h = 1, lty = 2, col = 'gray' )
```
